---
title: "lab_03"
author: "Derek Willis"
date: "2024-08-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse
library(tidyverse)

# Load janitor
library(janitor)
```

## Load data

```{r}
# Load MD slice of FEC House candidate contribution data
md_contribs <- read_rds("data/lab_03.rds")
```

## Fix header issues

You'll need to fix one issue with the column names using janitor. Write a code block to do that, making sure that you save the results to the same dataframe.

```{r}
md_contribs <- md_contribs |>
  clean_names ()

md_contribs
```


## Answer questions

Q1: Which seven cities in Maryland had the largest *number* of contributions from donors? How many did each have? To answer this question, you'll have to clean the messy "contributor_city" field.

Note: To answer this question, you'll need to see every variation of the "contributor_city" for the top seven cities, which you should identify through some initial analysis. Then find variations of them and change those variations into the most common version.

A1: Bethesda, Chevy Chase, Baltimore, Annapolis, Potomac, Silver Spring, and Rockville. Bethesda had 1177, Chevy Chase had 704, Baltimore had 690, Annapolis and Potomac had 490, Silver Spring had 405, and Rockville had 358. 

```{r}
top_contributions <-md_contribs |>
  group_by(contributor_city) |>
  summarise(total_contributions = n()) |>
  arrange(desc(total_contributions))

md_contribs <- md_contribs |>
  mutate(contributor_city = case_when(
    contributor_city == "Balrimore" ~ "Baltimore",
    contributor_city == "Blatimore" ~ "Baltimore",
     contributor_city == "Anapolis" ~ "Annapolis",
     contributor_city == "Potamic" ~ "Potomac",
     contributor_city == "Silver Springs" ~ "Silver Spring",
    TRUE ~ contributor_city
  )) 
  
  
```

Q2: Find the 5-digit ZIP code in Maryland that had the highest total amount of contributions (by dollar value). You will need to use mutate to ensure that you have a zip code column with the first five characters from the existing zip code column.

Where is that zip code and what kind of area is it, demographically? Which committee is the largest recipient of money from that zip code? How many of the top five recipients are Maryland candidates?

A2: *I used chatGPT to help figure out some parts of this code, particularly the summarise section* I believe the zip code is for Bethesda. Demographically, it is a majority white area with a median age of 43.5, according to World Population Review and Data USA. 

```{r}
md_contribs <- md_contribs %>%
  mutate (zip5 = substr(as.character(contributor_zip), 1, 5))

top_zip <- md_contribs  %>%
  group_by(zip5)  %>%
  summarise(total_contributions = sum(contribution_receipt_amount, na.rm = TRUE), .groups = "drop")  %>%
  arrange(desc(total_contributions))  %>%
  head(1)

print(top_zip)

```

Q3: Let's look at top individual donors. Write code to generate a list of top donors using name and city, adding up the total amount of contributions. What does the original data say about the top two donors - who are they and what do they do? Do you see anything in the list of top 10 donors that would make you uncertain about the accuracy of these results?

A3: *I also used ChatGPT to get some part of this code, also mainly in the summarise section* My results say that the top 2 contributors are Melanie Jacobsen and Mark Jacobsen from Bethesda. When I went back to look at the data, it says Mark's occupation is banking services, and also a CEO, with Promontory Interfinancial Group. It says Melanie is a framer at Clinton Farm. One thing that could potentially make me uncertain is that Mark Jacobsen is listed on there twice, at separate points. 

```{r}
top_donors <- md_contribs %>% 
  group_by(contributor_name, contributor_city) %>% 
  summarise(total_contributions = sum(contribution_receipt_amount, na.rm = TRUE), .groups = "drop") %>% 
  arrange(desc(total_contributions))

top_10_donors <- top_donors %>% head(10)
print(top_10_donors)

```
